Individual Variation in Animal Movement and Effects on Seed Dispersal Distance

### 2020-10-13 paper/paper.Rmd
changed methods, back to html again

abstract: |
  Seed dispersal is a fundamental process in plant ecology and due to its high importance in the ecosystem, various modelling approaches have been developed to understand seed movement and dispersal mechanisms. In particular, long-distance seed dispersal (LDD) plays a significant role in determining a population's genetic diversity and range expansion. Animal-mediated seed dispersal is a key component of LDD events, as frugivores can travel long distances and transport seeds with them. In this study we seek to understand the implications of individual variation in animal movement and how these can impact estimates of long distance seed dispersal. We apply these methods to empirical data sets that contain locations for an aracari, *Pteroglossus pluricintus*, one of the primary frugivores for the Amazonian canopy tree *Virola flexuosa*. We combine animal movement data and gut retention time to simulate *Virola* seed dispersal. This approach showed that there is significant variation in individual aracari movement patterns which directly influence the estimated number of long-distance dispersal events for *Virola* seeds. We found that models that ignore this underlying variation in individual animal movement, underestimate the percentage of long-distance seed dispersal events and maximum seed dispersal distances.
output: 
  word_document:
    reference_docx: paper_ref.docx
bibliography: mybibfile.bib
---

@@

In order to analyze the effect of differences in animal movement over final dispersal distances of foraged seeds we developed a spatially-explicit individual-based model in a homogeneous landscape that incorporated animal movement and seed gut retention time (GRT). Our simulation time steps were based on one minute intervals, and each simulation started at the origin or source tree, where each individual bird was given five seeds and was allowed to move in the landscape until the simulation ended once all seeds were dropped. Each seed's gut retention time was determined at the start of each simulation run, being randomly sampled from a gamma distribution (shape = 4, scale = 5, shift = 8) [@morales_effects_2006] so that the average gut retention time was 28 minutes[@holbrook_dispersal_2009; @holbrook_using_2007]. For each simulation run, the total simulation time was determined by the largest gut retention time sampled for the five seeds. We simulated animal movement using a memoryless process, a random walk with a uniformly distributed angular direction and a random step size chosen from a probability distribution. Commonly used probability distributions for sampling the animal movement step size included an exponential distribution [@jones_closing_2017], gamma distribution [citation], Weibull distribution[citation], and lognormal distribution [@levey_modelling_2008]. These four distributions are leptokurtic, and they vary in the fatness of their tails, thus allowing for different levels of variation in step size, with distributions such as the lognormal approaching a levy type walk (not sure this is all correct or makes sense).

>>>>

abstract: |
  Seed dispersal is a fundamental process in plant ecology and due to its high importance in the ecosystem, various modelling approaches have been developed to understand seed movement and dispersal mechanisms. In particular, long-distance seed dispersal (LDD) plays a significant role in determining a population's genetic diversity and range expansion. Animal-mediated seed dispersal is a key component of LDD events, as frugivores can travel long distances and transport seeds with them. In this study we seek to understand the implications of individual variation in animal movement and how these can impact estimates of long distance seed dispersal. We apply these methods to empirical data sets that contain locations for an aracari, *Pteroglossus pluricintus*, one of the primary frugivores for the Amazonian canopy tree *Virola flexuosa*. We combine animal movement data and gut retention time to simulate *Virola* seed dispersal. This approach showed that there is significant variation in individual aracari movement patterns which directly influence the estimated number of long-distance dispersal events for *Virola* seeds. We found that models that ignore this underlying variation in individual animal movement, underestimate the percentage of long-distance seed dispersal events and maximum seed dispersal distances.
output: html_document
bibliography: mybibfile.bib
---

@@

In order to analyze the effect of differences in animal movement over final dispersal distances of foraged seeds we developed a spatially-explicit individual-based model in a homogeneous landscape that incorporated animal movement and seed gut retention time (GRT). Our landscape consisted of a single source tree at the origin, our simulation time steps were based on one minute intervals, and each simulation started at the source tree (0,0), where each individual bird was given five seeds and was allowed to move freely in the landscape until the simulation ended once all seeds were dropped. Each seed's gut retention time was determined at the start of each simulation run, being randomly sampled from a Gamma distribution (shape = 4, scale = 5, shift = 8) [@morales_effects_2006] so that the average gut retention time was 28 minutes[@holbrook_dispersal_2009; @holbrook_using_2007]. For each simulation run, the total simulation time was determined by the largest gut retention time sampled for the five seeds, therefore, once the last seed was dropped, the simulation ended. From the simulation, we calculated the Euclidean distance from the source tree to each of the dropped seeds to determine the seed dispersal distance. We simulated animal movement using a memoryless process, a random walk with a uniformly distributed angular direction and a random step size chosen from a probability distribution. The simulation proceeded as follows: an individual bird starts the simulation at time zero and location zero zero. That bird is given five different seeds, with gut retention time (GRT_I), where GRT is gamma distributed. The animal is allowed to move freely across the landscape where at each time point T_i, a movement angle is sampled ~uniform(360) (think of greek letter) and a step size is randomly selected d_i, where D ~ f(x), where f(x) is the probability density function of one of the models we've selected. The movement process is repeated and at T_i == GRT_i a seed is dropped at the bird's location (x_i, y_i) and therefore that seed's final locations is (x_i, y_i). The simualation ends at T_i = max(GRT_1:5), and seed dispersal distances are calculated from the origin (0,0) to each seed's location (x_i, y_i).

# Simulations at different organizational levels

Commonly used probability distributions for sampling the animal movement step size included an exponential distribution [@jones_closing_2017], gamma distribution [citation], Weibull distribution[citation], and lognormal distribution [@levey_modelling_2008]. These four distributions are leptokurtic, and they vary in the fatness of their tails, thus allowing for different levels of variation in step size.
